---
title: "Longitudinal Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rphbPower)
```

## PART 1: REPEATED MEASURES ANALYSIS (for 2 timepoints)

Engine: `pwr::pwr.t.test` (type="paired") for robustness.

### Example 1.1: Sample Size for a Pre-Post Design

```{r ex1-1}
# Calculate N needed for 90% power to detect r=0.25 between two timepoints.
result_rm_1 <- repeated_measures_power(r_partial = 0.25, power = 0.90)
print(result_rm_1)
```

## PART 2: FIXED EFFECTS MODELS (for panel data)

Engine: Analytical t-test based on within-person degrees of freedom.

Use for panel data (many measurements from many units) to estimate within-person effects.

### Example 2.1: Sample Size for a Daily Diary Study

```{r ex2-1}
# Calculate the number of participants ('n_units') needed for a 14-day study,
# expecting a within-person effect of r=0.20 and an ICC of 0.4.
result_fe_1 <- fixed_effects_power(r_partial = 0.20, power = 0.8, n_periods = 14, icc = 0.4)
print(result_fe_1)
```

## PART 3: MIXED-EFFECTS (MULTILEVEL) MODELS

Engine: A robust Monte Carlo Simulation for Level-1 effects and an analytical engine for Level-2 effects. This section focuses on the Level-1 simulation, which supports two planning workflows.

### Workflow 1: Direct Replication (Default Behavior)

Use this approach when you have a **partial effect size** from a prior multilevel study and want to plan a direct replication. This is the default.

```{r ex3-1}
# Research Question: What is our power to replicate a within-person effect of r=0.25,
# given a study with 50 participants and 20 observations each?

# We provide the partial r from the original study. The function correctly uses
# inter_predictor_cor = 0 to avoid a "double penalty".
result_mm_1 <- mixed_models_power(
  r_partial = 0.25,
  n_groups = 50,
  n_per_group = 20,
  icc = 0.2,
  test_level = "level1"
)

print(result_mm_1)
```

### Workflow 2: Novel Study with Correlated Predictors

Use this approach when planning a new study from theory, starting with a simple (zero-order) effect size and an assumption about multicollinearity.


```{r 3.2}

# Research Question: We expect a simple effect of r=0.25, but we assume our
# within-person predictors will be correlated at r=0.3. What is our power
# with 50 participants and 20 observations each?

result_mm_2 <- mixed_models_power(
  r_partial = 0.25, # Our estimate of the simple effect
  n_groups = 50,
  n_per_group = 20,
  icc = 0.2,
  test_level = "level1",
  inter_predictor_cor = 0.3 # Explicitly model the multicollinearity
)

# Note: The power will be lower than in the Direct Replication workflow.
print(result_mm_2)

```


## PART 4: CROSS-LAGGED PANEL MODELS

Engine: Wraps linear_regression_power, using `pwr::pwr.f2.test` with n_predictors=2.

Use to test reciprocal effects (e.g., does X1 predict Y2, and Y1 predict X2?).

### Example 4.1: Sample Size for a Reciprocal Relationship
```{r ex4-1}
# Calculate the sample size needed to detect a cross-lagged path of r=0.15 with 80% power.
result_clpm_1 <- cross_lagged_panel_power(r_partial = 0.15, power = 0.8)
print(result_clpm_1)
```


## PART 5: DEMONSTRATING THE DISCOUNT FEATURE

Showcasing the conservative planning feature for a longitudinal study. Literature suggests a within-person effect of d=0.4 for an intervention.

```{r ex4}
# --- Conservative Planning (Recommended) ---
# Use the effect_input parameter to automatically apply the 0.75 discount.
conservative_result <- fixed_effects_power(
  effect_input = 0.4,
  effect_type = "d",
  power = 0.8,
  n_periods = 10,
  icc = 0.5
)

# --- Optimistic Planning (Not Recommended) ---
# Manually convert without the discount to see the difference.
optimistic_r <- cohens_d_to_partial_r(d = 0.4, apply_discount = FALSE)
optimistic_result <- fixed_effects_power(
  r_partial = optimistic_r,
  power = 0.8,
  n_periods = 10,
  icc = 0.5
)

cat(paste0(
  cat("\n--- Fixed Effects Conservative vs. Optimistic Planning ---\n"),
  cat("Literature Effect (Cohen's d):", 0.4, "\n"),
  cat("Conservative Plan (Framework Default): Requires N =", 
      conservative_result$n_units, "units\n"),
  cat("Optimistic Plan (No Discount): Requires N =", 
      optimistic_result$n_units, "units\n")
))
```

