---
title: "Longitudinal Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rphbPower)
```

## PART 1: REPEATED MEASURES ANALYSIS (for 2 timepoints)

Engine: `pwr::pwr.t.test` (type="paired") for robustness.

### Example 1.1: Sample Size for a Pre-Post Design

```{r ex1-1}
# Calculate N needed for 90% power to detect r=0.25 between two timepoints.
result_rm_1 <- repeated_measures_power(r_partial = 0.25, power = 0.90)
print(result_rm_1)
```

## PART 2: FIXED EFFECTS MODELS (for panel data)

Engine: Analytical t-test based on within-person degrees of freedom.

Use for panel data (many measurements from many units) to estimate within-person effects.

### Example 2.1: Sample Size for a Daily Diary Study

```{r ex2-1}
# Calculate the number of participants ('n_units') needed for a 14-day study,
# expecting a within-person effect of r=0.20 and an ICC of 0.4.
result_fe_1 <- fixed_effects_power(r_partial = 0.20, power = 0.8, n_periods = 14, icc = 0.4)
print(result_fe_1)
```

## PART 3: CROSS-LAGGED PANEL MODELS

Engine: Wraps linear_regression_power, using `pwr::pwr.f2.test` with n_predictors=2.

Use to test reciprocal effects (e.g., does X1 predict Y2, and Y1 predict X2?).

### Example 3.1: Sample Size for a Reciprocal Relationship
```{r ex3-1}
# Calculate the sample size needed to detect a cross-lagged path of r=0.15 with 80% power.
result_clpm_1 <- cross_lagged_panel_power(r_partial = 0.15, power = 0.8)
print(result_clpm_1)
```

## PART 4: DEMONSTRATING THE DISCOUNT FEATURE

Showcasing the conservative planning feature for a longitudinal study. Literature suggests a within-person effect of d=0.4 for an intervention.

```{r ex4}
# --- Conservative Planning (Recommended) ---
# Use the effect_input parameter to automatically apply the 0.75 discount.
conservative_result <- fixed_effects_power(
  effect_input = 0.4,
  effect_type = "d",
  power = 0.8,
  n_periods = 10,
  icc = 0.5
)

# --- Optimistic Planning (Not Recommended) ---
# Manually convert without the discount to see the difference.
optimistic_r <- cohens_d_to_partial_r(d = 0.4, apply_discount = FALSE)
optimistic_result <- fixed_effects_power(
  r_partial = optimistic_r,
  power = 0.8,
  n_periods = 10,
  icc = 0.5
)

cat(paste0(
  cat("\n--- Fixed Effects Conservative vs. Optimistic Planning ---\n"),
  cat("Literature Effect (Cohen's d):", 0.4, "\n"),
  cat("Conservative Plan (Framework Default): Requires N =", 
      conservative_result$n_units, "units\n"),
  cat("Optimistic Plan (No Discount): Requires N =", 
      optimistic_result$n_units, "units\n")
))
```

