---
title: "Mediation and SEM Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation and SEM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rphbPower)
```

Version: 2.0 - Revised to reflect validated engines and add context.

## PART 1: MEDIATION ANALYSIS (REGRESSION-BASED)

Engine: Analytical approximation based on the Aroian/Sobel test.

### Example 1.1: Sample Size Planning

```{r ex1-1}
# Target: 80% power to detect mediation where a-path r=0.20 and b-path r=0.30.
# Note: The function applies the 0.75 discount factor to these paths by default.
result_med_1 <- mediation_regression_power(r_a = 0.20, r_b = 0.30, power = 0.8)
print(result_med_1)
```

### Example 1.2: Demonstrating the Discount Feature

```{r ex1-2}
# Compare conservative vs. optimistic planning for pilot path coefficients.
pilot_a <- 0.45
pilot_b <- 0.35

# Conservative (recommended): The discount is applied to r_a and r_b before the calculation.
conservative_result <- mediation_regression_power(r_a = pilot_a, r_b = pilot_b, power = 0.8)

# Optimistic: Explicitly set the discount factor to 1.0 to disable it.
optimistic_result <- mediation_regression_power(r_a = pilot_a, r_b = pilot_b, power = 0.8, discount_factor = 1.0)

cat("\n--- Mediation Conservative vs. Optimistic Planning ---\n")
cat("Pilot Paths: a =", pilot_a, ", b =", pilot_b, "\n")
cat("Conservative Plan (Default, discount=0.75): Requires N =", conservative_result$n, "\n")
cat("Optimistic Plan (discount=1.0): Requires N =", optimistic_result$n, "\n")
```

## PART 2: MEDIATION ANALYSIS (SEM-BASED)

Engine: Analytical approximation, adjusted for measurement reliability.

NOTE: This is a fast approximation. For definitive results, a full Monte Carlo simulation is recommended.

### Example 2.1: Sample Size Planning with Reliability

```{r ex2-1}
# We expect latent paths of a=0.3 and b=0.25, with good measurement reliability (0.9).
result_sem_1 <- mediation_sem_power(r_a = 0.3, r_b = 0.25, power = 0.8, measurement_reliability = 0.9)
print(result_sem_1)
```

## PART 3: SEM DIRECT EFFECTS

Engine: t-test approximation for a single path, adjusted for reliability.

### Example 3.1: Sample Size for a Direct Path

```{r ex3-1}
# Target: 80% power to detect a direct path of r=0.20 in a model with 5 predictors,
# assuming measurement reliability of 0.85.
result_sem_dir_1 <- sem_direct_effects_power(r_partial = 0.20, power = 0.8, n_predictors = 5, measurement_reliability = 0.85)
print(result_sem_dir_1)
```

