---
title: "Specialized Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{specialized_models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rphbPower)
```

Version: 2.0 - Revised to reflect validated engines.

## PART 1: LOGISTIC REGRESSION

Engine: Monte Carlo Simulation for high accuracy with multiple predictors.

### Example 1.1: Power for a Binary Outcome

```{r ex1-1}
# Calculate the power to detect an odds ratio of 2.0 with N=400
# in a model with 3 total predictors.
result_log_1 <- logistic_regression_power(
  effect_input = 2.0,
  effect_type = "or",
  n = 400,
  n_predictors = 3
)
print(result_log_1)
```

# ==============================================================================
# PART 2: MIXED-EFFECTS MODELS
# ==============================================================================
# The engine for Level-1 effects is a robust Monte Carlo simulation.
# The engine for Level-2 effects is an analytical formula using `pwr::pwr.f2.test`.
# library(rphbPower)

# --- Example 2.1: Power for a Clustered Design (Level 1) ---
# Calculate the power to detect a within-group effect of r=0.25,
# given 50 groups with 25 individuals each and an ICC of 0.15.
# NOTE: The Level-1 simulation engine can only solve for power.
result_mix_1 <- mixed_models_power(
  r_partial = 0.25,
  n_groups = 50,
  n_per_group = 25,
  icc = 0.15,
  test_level = "level1"
)
print(result_mix_1)

### Example 2.2: Power for a Between-Group Effect (Level 2 Effect)

```{r ex2-2}
# Calculate power to detect an effect of r=0.30 between 50 schools.
result_mix_2 <- mixed_models_power(r_partial = 0.30, n_groups = 50, test_level = "level2")
print(result_mix_2)
```

## PART 3: WILCOXON SIGNED-RANK TEST

Engine: `pwr::pwr.t.test` (type="paired") as a robust parametric equivalent.

### Example 3.1: Sample Size for a Non-Normal Pre-Post Design

```{r ex3-1}
# Calculate N for 90% power to detect a medium effect (r=0.25).
result_wilcox_1 <- wilcoxon_signed_rank_power(r_partial = 0.25, power = 0.90)
print(result_wilcox_1)
```

## PART 4: DEMONSTRATING THE DISCOUNT FEATURE

Showcasing conservative planning for a nonparametric test. Literature suggests an effect of d=0.5.

```{r discount-feature}
# --- Conservative Planning (Recommended) ---
conservative_result <- wilcoxon_signed_rank_power(effect_input = 0.5, effect_type = "d", power = 0.8)

# --- Optimistic Planning (Not Recommended) ---
optimistic_r <- cohens_d_to_partial_r(d = 0.5, apply_discount = FALSE)
optimistic_result <- wilcoxon_signed_rank_power(r_partial = optimistic_r, power = 0.8)

cat("\n--- Wilcoxon Conservative vs. Optimistic Planning ---\n")
cat("Literature Effect (Cohen's d):", 0.5, "\n")
cat("Conservative Plan (Framework Default): Requires N =", conservative_result$n, "\n")
cat("Optimistic Plan (No Discount): Requires N =", optimistic_result$n, "\n")
```

